### 2021-07-06 rad-icfp-2021/generalization.tex
rewriting UNF section
fixing todo generalisation

Finally, the former point implies we need a few more array manipulations. 
We write !c::[v1,$\ldots$,vn]! to be the array ![c,v1,$\ldots$,vn]!. 
Given an array ![v1,$\ldots$,vn]! of size $n>0$, we write !let x,A=[v1,$\ldots$,vn]! to mean 
that !x=v1! and !A=[v2,$\ldots$,vn]!. The reverse derivative of fold left is then

TODO: reduce recomputed after scanl can be fixed here, explain let v,B notation somewhere
\begin{center}
\begin{tabular}{r c l}
    $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !foldl (x,y.e1) e2 e3!) &=&
            !let v,Y1 = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
            && !let A,Y2 = !$\directD{\rho}{\Gamma,v}{Y1}$!(e3) in! \\
            && !let A0 = (scanl (x,y.e1) v A)[1..n] in! \\
            && !let A1 = map2 (a,b.(!$\grad_{\{x\}}$!e1)[a/x,b/y]) A0 A! \\
            && !let A2 = map2 (a,b.(!$\grad_{\{y\}}$!)e1[a/x,b/y]) A0 A! \\
            && !let A3 = scanr * 1 A1! \\
            && !<reduce (x,y.e1) v A, fun (x1,$\ldots$,xn,z) ->! \\
            && !let y1,B = map2 (x,y. x*y*z) (1::A2) A3 in! \\
            && !Y2(x1,$\ldots$,xn,y1,B)>! \\
\end{tabular}
\end{center}

TODO: other examples/recipe

@@

With this definition, we recover that the gradient of $\Gamma \vdash$!e: !$\reals$ is a tuple of type $A1\times\ldotstimes An$ as expected.
TODO: define *, finish D(reduce)

>>>>

Finally, the former point implies we need a few more array manipulations.
These can be elegantly dealt with by changing the semantics of !scanl! and !scanr!. 
Let's assume that they now return a pair of an array of size $n$ 
of the intermediate computations and the final result.
The reverse derivative of fold left is then

\begin{center}
\begin{tabular}{r c l}
    $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !foldl (x,y.e$_1$) e$_2$ e$_3$!) &=&
            !let v,Y$_1$ = $\directD{\rho}{\Gamma}{Y}$(e$_2$) in! \\
            && !let A,Y$_2$ = $\directD{\rho}{\Gamma,v}{Y_1}$(e$_3$) in! \\
            && !let A$_0$,r$_1$ = (scanl (x,y.e$_1$) v A) in! \\
            && !let A$_1$ = map2 (a,b.(!$\grad_{\{x\}}$!e$_1$)[a/x,b/y]) A$_0$ A! \\
            && !let A$_2$ = map2 (a,b.(!$\grad_{\{y\}}$!)e$_1$[a/x,b/y]) A$_0$ A! \\
            && !let r$_2$, A$_3$ = scanr * 1 A$_1$! \\
            && !<r$_1$, fun (x$_1$,$\ldots$,x$_n$,z) ->! \\
            && !let y$_1$,B = r$_2$*z ,map2 (x,y. x*y*z) A$_2$ A$_3$ in! \\
            && !Y$_2$(x$_1$,$\ldots$,x$_n$,y$_1$,B)>! \\
\end{tabular}
\end{center}

Another useful example is !map!, which is very similar to !map2!.

\begin{center}
    \begin{tabular}{r c l}
        $\directD{\rho}{\Gamma}{Y}$(!map (x.e$_{1}$) e$_{2}$!) &=&  
            !let A,Y$_{1}$ = !$\directD{\rho}{\Gamma}{Y}$!(e$_{2}$) in! \\
            && !let G=!$\grad_{\Gamma}$!e$_{1}$ in!\\
            && !<map (x.e$_{1}$) A, fun (x$_{1}$,$\ldots$,x$_n$,Z) -> !\\
            && !Y$_{1}$( (x$_{1}$,$\ldots$,x$_n$)!$\widehat{+}$!(G*(reduce + 0 Z)),!\\
            && \quad\quad! map2 (a,b.(!$\grad_{\{x\}}$!e$_{1}$)[a/x]*b) A Z )>!\\
    \end{tabular}
\end{center}

@@

With this definition, we recover that the gradient of $\Gamma \vdash$!e: !$\reals$ is a tuple of type $A1\times\ldots\times An$ as expected.
TODO: define *, finish D(reduce)
